### 2016-02-11 discussion.Rmd
ENH:  Edits.
ENH:  More edits.

A recent meta-analysis by Kloppenborg et al. [@Kloppenborg:2014aa] of 23 cross-sectional studies reporting MRI and concurrent neuropsychological results in patients with heterogeneous diagnoses but without previously diagnosed cognitive impairment, found that WMHs were associated with cognitive deficit (effect size of -0.10, 95% CI: -0.13 to -0.08) after controlling for age. These studies also differed in the metric used to measure the WMHs, including volume, % of total intracranial volume, and a visual rating score. The effect size for the association with cognitive deficit in these cross-sectional studies did not differ significantly across various cognitive domains or the method of measuring lesion volume.  Among eight longitudinal studies analyzed by Kloppenborg et al that included a follow-up MRI and also controlled for age, the effect size for the association of progression in WMHs and cognitive impairment was -0.16 (95% CI:-0.27 to -0.09). This association was stronger for attention and executive function than for memory and processing speed. Although baseline WMHs were predictive of cognitive deficit at follow-up in the seven studies which did not repeat MRI, the effect size was smaller [-0.10 (95% CI: 0.13 to -0.05) than in the longitudinal studies that calculated progression in WMHs. In summary, progression of WMHs seen on repeat MRI has a stronger relation to cognitive deficit than concurrent imaging findings. These meta-analyses support the rationale for repeating an MRI in patients younger than 50 years whose initial scan shows WMHs.
Despite the above-described associations between WMHs, cognitive decline, increased risk of developing dementia, and mortality, these lesions receive little attention in current clinical workflows. When reported in a standard neuroradiologist interpretation, they are typically handled as incidental findings and are assigned little clinical significance. This likely reflects the impracticality of performing a detailed assessment of number, volume, and distribution within a qualitative neuroradiologist interpretation as well as the lack of correlative information on how the presence and distribution of these lesions may inform a diagnosis and prognosis in the appropriate clinical setting. To date, automated or semi-automated tools for the detection of WMHs have lacked the specificity and efficiency for the mining of large-scale datasets to generate highly granular data on whether these lesions possess any true diagnostic or prognostic value in the setting of a specific disease process. The present communication describes a supervised statistical learning tool that is appropriate for the application to such large-scale datasets.

The currently described tool is just one example of how “deep learning” algorithms might be applied to aid in the diagnosis of TBI and other disease processes through the specific identification of features predictive of a given disease state. It is an important demonstration of the potential power of these analytical approaches in the rapid but comprehensive mining of information from neuroimaging examinations. Deep learning algorithms are presently employed across a wide variety of settings for the rapid identification of predictive imaging features. Automobile manufacturers utilize these types of approaches to equip self-driving vehicles to recognize and respond to unique external surroundings through the identification of visual information sufficiently similar to previously assimilated training data. Similarly, in the context of the neuroimaging assessments, deep learning approaches may allow for the rapid identification of information predictive of disease state in an individual patient. These approaches may be applied to the identification of macroscopically visible lesions as described in the current communication. Additionally, these approaches might be applied to the interrogation of imaging data in the individual patient with a primary quantitative output metrics to include sequences such as diffusion tensor imaging (DTI) and its variants, functional connectivity, perfusion weighted imaging, and cortical thickness assessments. At present, these sequences are confined to cohort-based research studies due to the lack of available analytical tools to assess the information in the setting of the individual patient. Application of deep learning approaches in the context of data with primary quantitative outputs will require large scale normative and disease specific databases. Acquisition of these data is resource intensive and requires a multi-center approach with harmonized scanner between sites and correlative non-imaging clinical data. Large scale TBI data is becoming increasingly available through activities such as the Chronic Effects of Neurotrauma Consortium (CENC), Transforming Research and Clinical Knowledge in TBI (TRACK-TBI), Collaborative European Neurotrauma Effectiveness Research in TBI (CENTER-TBI), Department of Defense Alzheimer’s Disease Neuroimaging Initiative (DOD-ADNI), and other data being consolidated through the Federal Interagency TBI repository (FITBIR). In concert with any available high quality normative neuroimaging data, deep learning algorithms may be well positioned to help transform how neuroimaging is interpreted for the clinical management of patients with this disease process.

@@

>>>>

\replaced[id=dt]{The more}{A} recent meta-analysis by Kloppenborg et al. [@Kloppenborg:2014aa] of 23 cross-sectional studies reporting MRI and concurrent neuropsychological results in patients with heterogeneous diagnoses but without previously diagnosed cognitive impairment, found that WMHs were associated with cognitive deficit (effect size of -0.10, 95% CI: -0.13 to -0.08) after controlling for age. These studies also differed in the metric used to measure the WMHs, including volume, % of total intracranial volume, and a visual rating score. The effect size for the association with cognitive deficit in these cross-sectional studies did not differ significantly across various cognitive domains or the method of measuring lesion volume.  Among eight longitudinal studies analyzed by Kloppenborg et al that included a follow-up MRI and also controlled for age, the effect size for the association of progression in WMHs and cognitive impairment was -0.16 (95% CI:-0.27 to -0.09). This association was stronger for attention and executive function than for memory and processing speed. Although baseline WMHs were predictive of cognitive deficit at follow-up in the seven studies which did not repeat MRI, the effect size was smaller [-0.10 (95% CI: 0.13 to -0.05) than in the longitudinal studies that calculated progression in WMHs. In summary, progression of WMHs seen on repeat MRI has a stronger relation to cognitive deficit than concurrent imaging findings. These meta-analyses support the rationale for repeating an MRI in patients younger than 50 years whose initial scan shows WMHs.
Despite the above-described associations between WMHs, cognitive decline, increased risk of developing dementia, and mortality, these lesions receive little attention in current clinical workflows. When reported in a standard neuroradiologist interpretation, they are typically handled as incidental findings and are assigned little clinical significance. This likely reflects the impracticality of performing a detailed assessment of number, volume, and distribution within a qualitative neuroradiologist interpretation as well as the lack of correlative information on how the presence and distribution of these lesions may inform a diagnosis and prognosis in the appropriate clinical setting. To date, automated or semi-automated tools for the detection of WMHs have lacked the specificity and efficiency for the mining of large-scale datasets to generate highly granular data on whether these lesions possess any true diagnostic or prognostic value in the setting of a specific disease process. The present communication describes a supervised statistical learning tool that is appropriate for the application to such large-scale datasets.

The currently described tool is just one example of how "supervised learning" algorithms might be applied to aid in the diagnosis of TBI and other disease processes through the specific identification of features predictive of a given disease state. It is an important demonstration of the potential power of these analytical approaches in the rapid but comprehensive mining of information from neuroimaging examinations. Supervised learning algorithms are presently employed across a wide variety of settings for the rapid identification of predictive imaging features [@Plis:2014aa;@Suk:2015aa;@Li:2014aa;@Liu:2015aa]. Automobile manufacturers utilize these types of approaches to equip self-driving vehicles to recognize and respond to unique external surroundings through the identification of visual information sufficiently similar to previously assimilated training data [@HadsellSBESKML09;@FarabetCNL12]. Similarly, in the context of the neuroimaging assessments, deep learning approaches may allow for the rapid identification of information predictive of disease state in an individual patient. These approaches have been applied to the segmentation of macroscopically visible structures [@Plis:2014aa;@Suk:2015aa;@Li:2014aa;@Liu:2015aa]. Additionally, these approaches might be applied to the interrogation of imaging data in the individual patient with a primary quantitative output metrics to include sequences such as diffusion tensor imaging (DTI) and its variants, functional connectivity, perfusion weighted imaging, and cortical thickness assessments. At present, these advanced neuroimaging sequences are confined to cohort-based research studies due to the lack of available analytical tools to assess the information in the setting of the individual patient [@Mayer:2014aa]. Application of deep learning approaches in the context of data with primary quantitative outputs will require large scale normative and disease specific databases. Building these large scale imaging libraries is resource intensive and requires a multi-center approach with harmonized scanners between sites and correlative non-imaging clinical data. Large scale TBI data is becoming increasingly available through activities such as the Chronic Effects of Neurotrauma Consortium (CENC), Transforming Research and Clinical Knowledge in TBI (TRACK-TBI), Collaborative European Neurotrauma Effectiveness Research in TBI (CENTER-TBI), Department of Defense Alzheimer’s Disease Neuroimaging Initiative (DOD-ADNI), and other data being consolidated through FITBIR. In concert with any available high quality normative neuroimaging data, deep learning algorithms may be well positioned to help transform how neuroimaging is interpreted for the clinical management of patients with this disease process.

@@

<!--
\listofchanges

\clearpage
-->

### 2016-02-11 format.Rmd
ENH:  More edits.

---
output:
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
header-includes:
   - \usepackage{booktabs}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{1.5}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
bibliography: references.bib
csl: national-science-foundation-grant-proposals.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

>>>>

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
header-includes:
   - \usepackage{booktabs}
   - \usepackage[deletedmarkup=xout]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{1.5}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: national-science-foundation-grant-proposals.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

### 2016-02-11 intro.Rmd
ENH:  More edits.

@@

The random forests framework [@breiman2001] is a popular machine learning technique that has
demonstrated significant utitility for supervised segmentation tasks (e.g.,
normal human brain segmentation [@yi2009]) and other computer vision applications
(e.g., human gait detection [@viola2005]).
In the context of neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions [@geremia2011], stroke lesions [@Pustina:2016aa],
and brain tumors [@geremia2012;@bauer2012;@zikic2012;@Tustison:2015aa].  Of note, these
latter random forest approaches for brain tumor segmentation have performed well in recent
international competitions.  In response to the lack of objective comparisons between
segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in
2012 [@Menze:2015aa] and has continued every year since under the auspices of the MICCAI
conference.

Random forests are conceptually simple [@breiman2001].  They consist of ensembles of decision trees
that are built from training data.  Once constructed, data to be classified is "pushed"
through each decision tree resulting in a single classification "vote" per tree.  These
votes are then used for regression or classification of the data.  Although decision
trees had been extensively studied previously, the success of employing collections of
such weak learners for boosting machine learning performance
(e.g., AdaBoost [@schapire1990;@freund1997]) influenced the similarly sytled conglomeration
of decision trees into "forests" with randomized node optimization [@ho1995;@amit1997].
Finally, Breiman [@breiman2001] improved accuracy by random sampling of the training
data (i.e., "bagging") resulting in the current random forest technique.

In this work, we develop a concatenated random forest framework with a tailored contextual
feature image set (both spatial and intensity-based)
for segmenting white matter hyperintensities in traumatic brain injury cohorts.  Additionally,
the entire framework is provided publicly through the well-known open-source
ANTs[^i1] and ANTsR[^i2] toolkits.  Further motivating the research of this work is
the public availability of the imaging data thus permitting full reproducibility
of the results reported and discussed.

>>>>

<!--
A couple notes in response to comments:

[David Tate, Comment 15]:  Stupid question, but why not show the same subject and slice for both for a better comparison?

NT:  I see what he is saying and probably agree.

[David Tate, Comment 28]:  Maybe this could go in a Figure as a flow chart??

NT:  Flow charts are great and I included a couple as Figure 5 in my tumor paper.  However, they take a long time to make correctly and I do not want to commit to producing one unless it is absolutely necessary.

[Andy Mayer, Comment 30]: It will be difficult to see this font for labeling on Y axis at current size/resolution. You may also consider alerting your reader directly that the magnitudes of the scales are very different for stage 1 versus stage 2 in the legend or scale them to be similar.

NT:  Comparing these two plots in Figure 4 in terms of their absolute MeanDecreaseAccuracy
is meaningless so coordinating the scales is not important.

[Andy Mayer, Comment 34]:  A reviewer is likely to ask for some sort of reliability rating on the labeling of the lesions.

NT:  I agree with this but unless somebody else is willing to manually segment the WMHs,
then there is not really much we can do other than point to other papers which have looked
at reliability.

[David Tate, Comment 37]:  Seems like we are missing a picture.  That is the football player in me, but it seems like we need a picture that shows the manual compared to the automated at each stage if possible.

NT:  We could do this but I do not think that such images in publications are that meaningful as they typically result from heavy selection bias.  I did do something similar
for my brain segmentation paper where I showed multiple slices (skipped every 10, or something like that) for each subject in the challenge cohort but tumors are obviously much bigger and can be displayed in this manner.

[David Tate, Comment 39]:  I would would move these comments down to a limitation like paragraph, otherwise it minimizes the cool findings.

NT:  I have no opinion on this.  Do what you think is best.

[Andy Mayer, Comment 40]:  I do not know if you can include any of this data (even one subject). It would be great to demonstrate that your algorithm worked on data collected on another scanner since the intensities are likely to be very different.

NT:  As I mentioned to you previously, I think this is a great idea and I would be happy to include it but how do we showcase this unless somebody is willing to annotate data from these other sites.  We could include a picture of a single subject but how persuasive that is will be reviewer-dependent.  Personally, I would not be persuaded as the showing the results of a single subject is very prone to selection bias.

-->

@@

The random forests framework [@breiman2001] is a popular machine learning technique that has demonstrated significant \replaced[id=dt]{utility}{utitility} for supervised segmentation tasks (e.g.,
normal human brain segmentation [@yi2009]) and other computer vision applications
(e.g., human gait detection [@viola2005]).
In the context of neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions [@geremia2011], stroke lesions [@Pustina:2016aa],
and brain tumors [@geremia2012;@bauer2012;@zikic2012;@Tustison:2015aa].  Of note, these
latter random forest approaches for brain tumor segmentation have performed well in recent
international competitions \replaced[id=dt]{established in response to the lack of objective comparisons between segmentation algorithms (i.e.,  the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012}{
In response to the lack of objective comparisons between segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012 [@Menze:2015aa] and has continued every year since under the auspices of the MICCAI conference.}[@Menze:2015aa].

Random forests are conceptually simple [@breiman2001].  They consist of ensembles of decision trees
that are built from training data.  Once constructed, data to be classified is "pushed"
through each decision tree resulting in a single classification "vote" per tree.  These
votes are then used for regression or classification of the data.  Although decision
trees had been extensively studied \deleted[id=dt]{previously}, the success of employing collections of
such weak learners for boosting machine learning performance
(e.g., AdaBoost [@schapire1990;@freund1997]) influenced the similarly sytled conglomeration
of decision trees into "forests" with randomized node optimization [@ho1995;@amit1997].
Finally, Breiman [@breiman2001] improved accuracy by random sampling of \deleted[id=dt]{the} training
data (i.e., "bagging") resulting in the current random forest technique \added[id=dt]{applied here}.

In this work, we develop a concatenated random forest framework with a tailored contextual
feature image set (both spatial and intensity-based)
for segmenting \replaced[id=dt]{WMHs}{white matter hyperintensities} in \replaced[id=dt]{a arge TBI cohort}{traumatic brain injury cohorts}.  Additionally,
the entire framework is provided publicly through the well-known open-source
\added[id=lw]{Advanced Normalization Tools} (ANTs)[^i1] and ANTsR[^i2] toolkits.
\replaced[id=dt]{Further motivating this research is the availability of several large publicly available  imaging data sets that permits testing reproducibility of this automate routine for WMH segmentation and quantification.}{Further motivating this research of this work is
the public availability of the imaging data thus permitting full reproducibility
of the results reported and discussed.}

### 2016-02-11 methods.Rmd
ENH:  More edits.

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.  In Table 1, we provide a list and basic
categorization of the feature images used for the initial (i.e., Stage 1---more on the use
of multiple random forest stages below) segmentation of the white matter hyperintensities.
In addition Figure 1 provides a representation of a set of feature images for a single
subject analyzed in this work.  Note that in this work we categorize the brain parenchyma
with seven labels:

@@

As mentioned previously, input for each subject comprises FLAIR, T1-, and T2-weighted
acquisitions.  The FLAIR and T2 images are rigidly registered to the T1 image using
the open-source Advanced Normalization Tools (ANTs) [@Avants:2014aa].  The aligned images
are then preprocessed using the denoising algorithm of [@Manjon:2010aa] followed by
N4 bias correction [@Tustison:2010ac] which are then normalized to the intensity
range $[0,1]$.  Although we could have used an alternative intensity standardization
algorithm (e.g., [@nyul2000]), we found that a simple linear rescaling produced better results.

@@

The T1 image is then processed via the ANTs brain
extraction and normal tissue segmentation (i.e., the pathology is discounted which is
a fairly safe assumption for the T1 image) protocols as part of the
cortical thickness pipeline described in [@Tustison:2014ab].
\added[id=hl]{This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is} a mask delineating the brain parenchyma and probabilistic estimates of the
cerebrospinal fluid (csf), gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  These provide the expertly annotated labels for the first six tissue labels
given above.  The white matter hyperintensities were manually identified by one of the authors
(J. R. S.) using the ITK-SNAP tool [@Yushkevich:2006aa].  Segmentation is performed using
the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2015aa] created from the public
MMRR data set  [@landman2011] (cf Figure 3).

@@

To model the intensity information the first set of images simply includes the
preprocessed and normalized intensity FLAIR, T1, and T2 image voxel values.  We also calculate a set
of neighborhood statistics (mean, standard deviation, and skewness) feature images using
a \added[id=am]{Manhattan} radius of one voxel. For each of the normalized
images, we calculate the difference in intensities with the corresponding warped template
component.  Previous success in the international brain tumor segmentation
competition [@Menze:2015aa] was based on an important set of intensity features that were created
from multi-modal templates mentioned previously [@Tustison:2015aa].  We employ the
same strategy here.  For example, the template
difference feature image for the FLAIR image, $S_{FLAIR}$ is calculated as:

@@

where $\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$ is the transform
which maps from the individual subject space to the
template space and $T_{FLAIR}$ is the FLAIR template component.  Also,
to take advantage of the \added[id=hl,remark=NT:  we are using the "symmetry" of the brain in an analogous sense that if we know the left eye location and color then it is highly probabilistic that we can accurately predict the location and color of the right eye even though a particular face might not be exactly symmetric]{gross} bilateral symmetry of the normal brain (in terms of both shape and intensity), and the fact that \replaced[id=bt]{WMHs do not generally manifest symmetrically across hemispheres}{the presence of WMHs violates that assumption},
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature.  For the FLAIR component, this contralateral
difference image is calculated from

@@

The segmentation probability images described above are used as feature images to provide a
spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.  The T1 image of the subject is registered to the T1 template
component using a B-spline variant [@Tustison:2013ac] of the well-known ANTs Symmetric
Normalization (SyN) algorithm [@Avants:2011ab].  Since the inverse transform is also
derived as part of the registration process, we can warp the voxel index locations
back to the space of the individual subject.    Note that this is similar in motivation to the work
of [@Anbeek:2004aa].  However, this previous work lacks the normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion segmentation review
[@Garcia-Lorenzo:2013aa], although the number of algorithms reported in the literature
is quite extensive, there were only four publicly available segmentation algorithms
at the time of writing of which none are based on supervised learning.  As we did for
our brain tumor segmentation algorithm [@Tustison:2015aa], all of the code described in
this work is publicly available through the open-source ANTs/ANTsR toolkits.  Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R statistical project) we
use the \texttt{randomForest} package [@liaw2002] using the default settings with
2000 trees per model and 500 randomly selected samples per label per image.
\added[id=bt]{Note that we saw little variation in performance when these parameters were changed which is consistent with our previous experience.}

In addition, similar to our previous offering [^1], we plan on creating
a self-encapsulated example to showcase the proposed methodology.  The fact that
the data will also be made available through the FITBIR repository along
with the manual labelings will facilitate reproducibility on the part of the reader
as well as any interest in extending the proposed framework to other data sets.

>>>>

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.  In Table 1, we provide a list and basic
categorization of the feature images used for the initial (i.e., Stage 1---more on the use
of multiple random forest stages below) segmentation of the WMHs.
In addition Figure 1 provides a representation of a set of feature images for a single
subject analyzed in this work.  Note that in this work we categorize the brain parenchyma
with seven labels:

@@

As mentioned previously, input for each subject comprises FLAIR, T1-, and T2-weighted
acquisitions.  The FLAIR and T2 images are rigidly registered to the T1 image using
the open-source Advanced Normalization Tools (ANTs) [@Avants:2014aa].  The aligned images
are then preprocessed using the denoising algorithm of [@Manjon:2010aa] followed by
N4 bias correction [@Tustison:2010ac] which are then normalized to the intensity
range $[0,1]$.  Although we could have used an alternative intensity standardization
algorithm (e.g., [@nyul2000]), we found that a simple linear rescaling produced better results similar \added[id=dt]{to previous work}[@Tustison:2015aa].

@@

The T1 image is then processed via the ANTs brain
extraction and normal tissue segmentation (i.e., the pathology is discounted which is
a fairly safe assumption for the T1 image) protocols as part of the
cortical thickness pipeline described in [@Tustison:2014ab].
\added[id=hl]{This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is} a mask delineating the brain parenchyma and probabilistic estimates of the
\replaced[id=dt]{CSF}{cerebrospinal fluid (csf)}, gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  These provide the expertly annotated labels for the first six tissue labels
given above.  The \replaced[id=dt]{WMHs}{white matter hyperintensities} were manually identified by one of the authors
(J. R. S.) using the ITK-SNAP tool [@Yushkevich:2006aa].  Segmentation is performed using
the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2015aa] created from the public
MMRR data set  [@landman2011] (cf Figure 3).

@@

To model the intensity information the first set of images simply includes the
preprocessed and normalized intensity FLAIR, T1, and T2 image voxel values.  We also calculate a set
of neighborhood statistics (mean, standard deviation, and skewness) feature images using
a \added[id=am]{Manhattan} radius of one voxel \added[id=nt]{given the typical size
of individual WMHs}. For each of the normalized
images, we calculate the difference in intensities with the corresponding warped template
component.  Previous success in the international brain tumor segmentation
competition [@Menze:2015aa] was based on an important set of intensity features that were created
from multi-modal templates mentioned previously [@Tustison:2015aa] \added[id=nt]{and listed in Table 1}.  We employ the
same strategy here.  For example, the template
difference feature image for the FLAIR image, $S_{FLAIR}$ is calculated as:

@@

where $\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$ is the transform
which maps from the individual subject space to the
template space and $T_{FLAIR}$ is the FLAIR template component.  Also,
to take advantage of the \added[id=hl,remark=NT:  we are using the gross bilateral symmetry of the brain in an analogous sense that if we know the left eye location and color then we can accurately predict the location and color of the right eye even though a particular face might not be exactly symmetric]{gross} bilateral symmetry of the normal brain (in terms of both shape and intensity), and the fact that \replaced[id=bt]{WMHs do not generally manifest symmetrically across hemispheres}{the presence of WMHs violates that assumption},
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature.  For the FLAIR component, this contralateral
difference image is calculated from

@@

The segmentation probability images described above are used as feature images to provide a
spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.  The T1 image of the subject is registered to the T1 template
component using a B-spline variant [@Tustison:2013ac] of the well-known ANTs Symmetric
Normalization (SyN) algorithm [@Avants:2011ab].  Since the inverse transform is also
derived as part of the registration process, we can warp the voxel index locations
back to the space of the individual subject \replaced[id=nt]{which motivates similar work by others}{
Note that this is similar in motivation to the work
of} [@Anbeek:2004aa].  However, this previous work lacks the normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion segmentation review
[@Garcia-Lorenzo:2013aa], although the number of algorithms reported in the literature
is quite extensive, there were only four publicly available segmentation algorithms
at the time of writing \added[id=am]{this article}.  \replaced[id=dt]{In contrast to the current work,}{of which} none are based on supervised learning.  As we did for
our brain tumor segmentation algorithm [@Tustison:2015aa], all of the code described in
this work is publicly available through the open-source ANTs/ANTsR toolkits.  Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R statistical project) we
use the \texttt{randomForest} package [@liaw2002] using the default settings with
2000 trees per model and 500 randomly selected samples per label per image.
\added[id=bt]{Note that we saw little variation in performance when these parameters were changed which is consistent with our previous experience.}

In addition, similar to our previous offering [^1], we plan on creating
a self-encapsulated example to showcase the proposed methodology.  The fact that
the data will also be made available through the \replaced[id=lw]{Federal Interagency Traumatic Brain Injury Research (FITBIR)}{FITBIR} repository along
with the manual labelings will facilitate reproducibility on the part of the reader
as well as any interest in extending the proposed framework to other data sets.

### 2016-02-11 results.Rmd
ENH:  More edits.

After performing the leave-one-out evaluation described at the end of the previous section,
we calculated the \texttt{MeanDecreaseAccuracy} feature values for each of the 24 subjects $\times$
2 models per subject $=48$ total models.  This measure (per feature, per model) is calculated
during the out-of-bag phase of the random forest model construction and quantifies the decrease
in prediction accuracy from omitting the specified feature.  In other words, this quantity
helps determine the importance of a particular feature and, although we save such
efforts for future work, this information provides us with guidance for future feature
pruning and/or additions.

@@

![Comparison with manual delineation of white matter hyperintensities.  On the left are the
calculate Dice values over all white matter hyperintensities.  Note the improvement
in the Dice metric from the employment of the Stage 2 component of the processing pipeline.
(Right) Similar results can be seen by comparing the total lesion load volume between manual
and automated detection strategies.  Although some outliers are found after the Stage 2
processing in a couple subjects, the number of outliers caused by false positives is
significantly with the second stage processing.](Figures/llvAndDice.png)

>>>>

After performing the leave-one-out evaluation\deleted[id=dt]{described at the end of the previous section},
we calculated the \texttt{MeanDecreaseAccuracy} feature values for each of the 24 subjects $\times$
2 models per subject $=48$ total models.  This measure (per feature, per model) is calculated
during the out-of-bag phase of the random forest model construction and quantifies the decrease
in prediction accuracy from omitting the specified feature.  In other words, this quantity
helps determine the importance of a particular feature and, although we save such
efforts for future work, this information provides us with guidance for future feature
pruning and/or additions.

@@

![\added[id=am]{Voxelwise} comparison with manual delineation of white matter hyperintensities.  On the left are the
calculate\added[id=am]{d} Dice values over all white matter hyperintensities.  Note the improvement
in the Dice metric from the employment of the Stage 2 component of the processing pipeline.
(Right) Similar results can be seen by comparing the total lesion load volume between manual
and automated detection strategies.  Although some outliers are found after the Stage 2
processing in a couple subjects, the number of outliers caused by false positives is
\added[id=lw]{decreased} significantly with the second stage processing.](Figures/llvAndDice.png)
